Shared Subspace Model

### 2016-06-19 paper/mscov.tex
Intro edits for Peter

\documentclass{article}

\usepackage{graphicx}
\usepackage{latexsym}
\usepackage{textcomp}
\usepackage{longtable}
\usepackage{multirow, booktabs}
\usepackage{amsfonts,amsmath,amssymb}
\usepackage{natbib}
\usepackage{url}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{chemarrow}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{tikz}
\usepackage[margin=.75in]{geometry}

@@

  We develop a model-based method for evaluating similarities and
  differences among several $p\times p$ covariance matrices in the
  $p > n$ setting.  This is done by assuming a spiked covariance model
  for each group, and sharing information about the subspace spanned
  by the group-level eigenvectors.  We propose an EM algorithm for
  estimating the shared subspace and the covariance matrix of the data
  projected onto this subspace.  We also provide a Gibbs sampling
  algorithm for estimating the posterior uncertainty of group-level
  eigenvectors and eigenvalues.  Finally, we demonstrate the utility
  of our method in an analysis of biological data.

@@

It has long been known that the sample covariance matrix is a poor
estimator of the population covariance matrix, particularly when the
sample size is comparable to or smaller than the number of features
($n \leq p$) \citep{Dempster1969, Stein1975}.  A host of estimators
have been devised to address this fact. These primarily include of
shrinkage based approaches \citep{Schafer2005, Ledoit2011} and low
rank factor models \citep{Mardia1980, Ullman2003, Fan2008}.  In many
cases, the covariance matrices are highly structured and thus can be
sparsely parameterized even if the matrix itself is not low rank
\citep{Burg1982, Williams1993} .  Along these lines, there has been a
recent interest in sparse Gaussian graphical model, in which the there
are assumed to be many zeros in the inverse covariance matrix
\citep{Meinshausen2006, Friedman2008}.

Most of these approaches assume that the goal is to estimate a
covariance matrix from a single dataset.  In many applications,
however, the data can be split into groups, each of which represent
samples from populations with distinct (but possibly related)
covariance matrices.  In this case, a comparative analysis of
covariance patterns across multiple groups of observations is often
warranted.  Evaluating similarities and differences between covariance
matrices can be an important complement to classical analyses which
center primarily on estimating differences in means across groups.
For instance, in case control-studies, mean level effects may be small
relative to subject variability; distributional differences between
groups may still be evident as differences in the covariances between
features.  Even when mean level differences are detectable, better
estimates of the covariability of features across groups may lead to
an improved understanding of the mechanisms underlying these apparent
mean level differences.  Further, accurate covariance estimation is an
essential part of many prediction tasks (e.g. quadratic discriminant
analysis).

By exploiting similarity across related groups of high-dimensional
data, we can increase the efficiency of existing covariance
estimators.  Some of the earliest approaches for modeling multiple
covariance matrices focused on estimation in terms of the spectral
decompositions for these matrices.  \cite{Flury1987} developed
estimation and testing procedures for the ``common principal
components'' model, in which a set of covariance matrices were assumed
to share the same eigenvectors.  \citet{Schott1991, Schott1999}
considered cases in which only certain eigenvectors are shared across
populations, and \citet{Boik2002} described an even more general model
in which eigenvectors can be shared between all or some of the groups.
More recently, \citet{Hoff2009}, noting that eigenvectors are unlikely
to be shared exactly between groups, introduced a hierarchical model
for eigenvector shrinkage based on the matrix Bingham distribution.
There has also been a significant interest in estimating covariance
matrices using Gaussian graphical models. For Gaussian graphical
models, zeros in the precision matrix correspond to conditional
independences between pairs of features given the remaining features
\citep{Meinshausen2006}.  \citet{Witten2014} extended existing work in
this area to the multi-group setting, by pooling information about the
pattern of zeros in across precision matrices.

@@

Another popular method for modeling relationships between
high-dimensional multivariate data related that is closely related to
low rank covariance estimation, is known partial least squares
regression (PLS). This approach, which is a special case of a bilinear
factor model, involves projecting the data into a lower dimension
space which maximizes the similarity of the two groups; this technique
does not require the data from each group to share the same feature
set.  An oft used variant for prediction, partial least squares discriminant analysis
(PLS-DA) is especially common in the chemometrics and bioinformatics
 \citep{Barker2003} .  Although closely related to
the approaches we will consider here, the primarily goal of of PLS based models
is to create regression or discrimination models, not to explicitly
infer covariance matrices from multiple groups of data.  Nevertheless,
the basic idea that data can often be well represented on a low
dimensional space is an appealing one that we leverage as well.

Motivated by this notion, in this paper we propose a multi-group
covariance estimation model by sharing information about the subspace
spanned by group-level eigenvectors.  The shared subspace assumption
can be used to improve estimates and facilitate the interpretation of
differences between covariance matrices across groups.  For each
group, we assume ``the spiked covariance model'', a well studied
variant of the factor model \citep{Johnstone2001}.  In Section
\ref{sec:shared} we briefly review the behavior of spiked covariance
models for estimating a single covariance matrix before introducing
our extension to the multi-group setting.  In Section
\ref{sec:inference} we describe an optimization algorithm for
inferring the shared subspace and then outline a MCMC algorithm for
estimating the posterior distribution of the projected covariance
matrices.  Finally, in Section \ref{sec:simulation} we investigate the
behavior of these models in simulation before demonstrating the
utility of this model with real biological data in Section
\ref{sec:app}.

@@

\paragraph{The Spiked Covariance Model:}

Suppose $\bl S$ has a possibly degenerate Wishart$(\Sigma,n$)
distribution with density given by
\begin{equation} 
p(\bl S | \Sigma, n) \propto_{\Sigma} l(\Sigma:\bl S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1} \bl S/2 ) ,  
\label{eqn:lik}
\end{equation}
%
\noindent where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.  Such
a likelihood results from $\bl S$ being, for example, a residual sum
of squares matrix from a multivariate regression analysis. In this
case, $n$ is the number of independent observations minus the rank of
the design matrix.  The spiked principle components model (spiked PCA)
introduced by \citet{Johnstone2001} assumes that
\begin{equation} 
\Sigma = \sigma^2 (  \bl U  \Lambda  \bl U^T  + \bl I )
\label{eqn:spiked}
\end{equation}
%
\noindent where $\Lambda$ is an $r\times r$ diagonal matrix and
$\bl U \in \mathcal V_{r,p}$, with $r \ll p$. Here $\mathcal V_{r,p}$,
known as the Stiefel manifold, comprises the set of $p \times r$
orthonormal matrices in $\mathcal{R}^p$.  The spiked covariance model
is thus a low rank factor model with homoscedastic error.  Such a
formulation is appealing because it explicitly partitions the
covariance matrix into a tractable low rank ``signal'' and isotropic
``noise''.

@@

Classical results for parametric models (e.g., \citet{Schwartz1965})
imply that asymptotically in $n$ for fixed $p$, a Bayes estimator will
be consistent for a spiked population covariance as long as the
assumed number of spikes (non-unit eigenvalues) is greater than or
equal to the true number.  However, when $p$ is large relative to $n$,
as is the case for the examples considered here, things are more
difficult.  Under the spiked covariance model, it has been shown that
$p/n \rightarrow \gamma >0$ as $n\rightarrow \infty$, the top $r$
eigenvalues of $\bl S/(n\sigma^2)$ will converge to upwardly biased
versions of $(\lambda_1+1,\ldots, \lambda_r+1)$, if each $\lambda_s $
is greater than $\sqrt{\gamma}$ \citep{Baik2006, Paul2007} .  This has
led several authors to suggest estimating $\Sigma$ via shrinkage of
the eigenvalues of the sample covariance matrix. In particular, in the
setting that $\sigma^2$ is known, \citet{Donoho2013} propose setting
all sample eigenvalues that fall below the value
$\sigma^2(1+\sqrt{\gamma})^2$ to be $\sigma^2$, and shrinking the
larger eigenvalues in a way that depends on the particular loss
function being used.  These shrinkage functions are shown to be
asymptotically optimal in the $p/n\rightarrow \gamma$ setting.

@@

Many eigenvalue shrinkage estimators can be viewed in the Bayesian
context as derivative of a Bayes estimator under a particular prior.
Note that in the absence of real prior information, invariance
considerations suggest the use of priors on $\sigma^2$ and $\bl U$
that lead to equivariant estimators.  For instance, an (improper)
prior having density $p(\sigma^2,\bl U) = 1/\sigma^2 $, with respect
to the product of Lebesgue measure on $\mathbb R^+$ and the uniform
probability measure on $\mathcal V_{r,p}$, leads to Bayes estimators
$\hat \Sigma(\bl S)$ that are equivariant with respect to rotations
and scale changes.  For multi-group covariance estimation, however, we
often have reason to believe that a priori the eigenvectors are
related across groups.  In this case, group-level equivariance to
rotations is an unreasonable assumption.  Here, both eigenvalue
\emph{and} eigenvector shrinkage can play an important role in
improving covariance estimates.  Below, we introduce our extension of
the spiked covariance model to the group setting by including an
assumption about the space shared by the span of the group-level eigenvectors.

\paragraph{The Shared Subspace Model:}

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/3dplot}
        \caption{Projection in $\mathcal{R}^3$}
        \label{fig:dmelanRatio}
    \end{subfigure}
\quad
 %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter}
        \caption{$Y_kV$}
        \label{fig:dmelanPosterior}
    \end{subfigure}
\quad 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter-orth}
        \caption{$Y_kV_{\perp}$}
        \label{fig:dmelanPosterior}
    \end{subfigure}
    \caption{ Two groups of four-dimensional data (red and blue)
      projected into different subspaces.  a) To visualize $Y_k$ we
      can project the data into $R^3$.  In this illustration, the
      distributional differences between the groups are confined to a
      two-dimensional shared subspace ($VV^T$, grey plane).  b) The
      data projected onto the two-dimensional shared subspace,
      $Y_kV$, have covariances $\psi_k$ that differ between
      groups. c) The orthogonal projection, $Y_kV_{\perp}$
      has isotropic covariance, $\sigma_k^2I$, for all groups.  }
\label{fig:shared}
\end{figure}

The spiked covariance model assumes that the goal is to estimate a
single covariance matrix.  The situation is different when we are
interested in estimating multiple covariance matrices from distinct,
but related groups.  For multi-group estimation, we can often improve
upon the efficiency of covariance estimators by pooling information
across the groups of data.  Here, we assume we have $k=1, ..., K$
groups of mean-zero normal data, $Y_k$, where $Y_k$ consists of $n_k$
independent rows of measurements on $p$ features, typically with
$n_k \ll p$.  Then, $Y_k^TY_k = \bl S_k$ follow a (degenerate) Wishart
distribution as in Equation \ref{eqn:lik}.  To improve estimation, we
seek estimators for the population covariance from one group that depend on the
data from all $K$ groups, e.g. of the form
$\hat \Sigma_k(\bl S_1, ... \bl S_K)$.  To this end, we extend the
spiked covariance model to grouped data, by assuming that the
anisotropic variability from each group can be expressed on a common
low dimensional subspace.  Specifically, we posit that the covariance
matrix for each group can be written as

\begin{equation}
\Sigma_k = V\psi_kV^T + \sigma^2_kI
\label{eqn:sspsi}
\end{equation}
%
\noindent Here, the columns of $V \in \mathcal{V}_{p, s}$ determine a subspace
shared by all groups $k$ with $p \geq s$.  Throughout this paper we
will denote to the shared subspace as $VV^T \in \mathcal G_{r,p}$, where
$\mathcal G_{r,p}$, known as the Grassmann manifold, consists of all
$s$-dimensional linear subspaces of $\mathcal{R}^p$
\citep{Chikuse2012}.  Although $V$ is only identifiable up to right
rotations, $VV^T$, which defines the plane of variation shared by all
groups, is identifiable.  Later, to emphasize the connection to the
spiked PCA model (\ref{eqn:spiked}), we will write $\psi_k$ in terms
of its eigendecomposition, $\psi_k = O_k\Lambda_kO_k$, where $O_k$ are
eigenvectors and $\Lambda_k$ are the eigenvalues of $\psi_k$ (see
Section \ref{sec:bayes}).

Note that for the shared subspace model,
$V^T\Sigma_kV = \psi_k + \sigma_k^2I$ is an anisotropic
$s$-dimensional covariance matrix for the projected data, $Y_kV$.
In contrast, the data projected onto the orthogonal space,
$Y_kV_{\perp}$, is isotropic, and typically small, for all
groups.  In Figure \ref{fig:shared} we provide a simple illustration
of this fact using simulated $4$-dimensional data from two groups.  The
differences in distribution between the groups of data can be
expressed on a two dimensional subspace spanned by the columns of
$V \in \mathcal{V}_{4, 2}$.  Differences in the correlations between
the two groups manifest themselves on this shared subspace, whereas
only the magnitude of the isotropic variability can differ between groups
on the orthogonal space.  Thus, the the shared subspace model can be
viewed as a covariance partition model, where one partition includes
the anisotropic variability from all groups and the other partition is
constrained to the isotropic variability from each group.  This
isotropic variability is often taken to be ``measurement noise''.

@@

\section{Inference}
\label{sec:inference}
In this section we outline Bayesian inference algorithms including
both an optimization algorithm and an MCMC algorithm for estimating
covariance parameters and their associated uncertainty.  MCMC
inference is particularly appealing because of the ease with which we
can quantify the uncertainty associated with each estimate.
Uncertainty quantification is particularly important in the $n \ll p$
setting where seemingly large differences in the point estimates of
the low rank covariance matrices across groups may not actually
reflect significant differences.  Although uncertainty quantification
for all parameters is desirable, for the size of the problems we are
generally interested in, we found MCMC inference for the subspace
$VV^T$ to be intractable.  Our optimization based approached is
significantly faster and works well for high-dimensional datasets.  As
such, we start, in Section \ref{sec:em}, by outlining an
expectation-maximization algorithm for estimating the shared
subspace. In Section \ref{sec:bayes} we describe a Gibbs sampler for
posterior inference of $\psi_k$ given the common subspace.  These two
algorithms can be used in conjunction to first identify and evaluate a
shared subspace and then to quantify the uncertainty associated with
the projected covariance matrices.

@@

In this section we describe a computationally efficient algorithm for
estimating the shared subspace, $VV^T$, the projected covariance matrix
$\psi_k$ and noise parameter $\sigma_k^2$.  First, note that the full
likelihood for the shared subspace model can be written as

\begin{align}
p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}\mathbf{S}_k/2)  \\
&\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\psi_kV^T +
  I))^{-1}\mathbf{S}_k/2) \\
&\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]\mathbf{S}_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]\mathbf{S}_k/2) 
\end{align}
%
\noindent where we define $M_k^{-1} = (\psi_k + I) ^{-1}/\sigma_k^2$.  The log-likelihood in
$V$ (up to additive constant) is

\begin{align}
l(V) &= \sum_k tr\left(-VM_k^{-1}V^T +
       VV^T/\sigma^2_k\right)\mathbf{S}_k/2)\\
&= \sum_k \frac{1}{2}tr\left(-M_k^{-1}V^T \mathbf{S}_kV\right) + \frac{1}{2\sigma_k^2}tr\left(VV^T \mathbf{S}_k\right)
\end{align}

@@

\noindent which is a Wishart($V^T\mathbf{S}_kV$, $n_k$).  The
conditional posterior distribution of $\frac{1}{\sigma_k^2}$ is simply

@@

\noindent which is a Gamma($n_k(p-r)/2$,
$\frac{1}{2}tr((I-VV^T)\mathbf{S}_k)$).  Thus, for each iteration,
$t$, of the EM algorithm:

\begin{enumerate}
\item For each $k$, compute relevant conditional expectations:
\begin{enumerate}
\item $E[M_k^{-1} | V_{(t-1)}] = n_k(V_{(t-1)}^T \mathbf{S}_kV_{(t-1)})^{-1}$
\item $E[\frac{1}{2\sigma_k^2}|V_{(t-1)}] = \frac{n_k(p-r)}{tr((I-V_{(t-1)}V_{(t-1)}^T)S_k)}$
\end{enumerate}
\item Compute $V_{t}$ = $\underset{V}{argmax}  \sum_k tr\left(-VE[M^{-1}|V_{(t-1)}]^T +
       E[\frac{1}{2\sigma_k^2}|V_{(t-1)}]VV^T\right)\mathbf{S}_k/2)$ 
\end{enumerate}

@@

\paragraph{Evaluating a Shared Subspace:}

>>>>

\documentclass[12pt]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{latexsym}
\usepackage{textcomp}
\usepackage{longtable}
\usepackage{multirow, booktabs}
\usepackage{natbib}
\usepackage{url}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{chemarrow}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{tikz}
\usepackage[margin=.75in]{geometry}

@@

  We develop a model-based method for evaluating similarities and
  differences among several $p\times p$ covariance matrices in the
  $p \gg n$ setting.  This is done by assuming a spiked covariance
  model for each group and sharing information about the space spanned
  by the group-level eigenvectors.  We use empirical Bayes to identify
  a low-dimensional subspace which explains as a large amount of
  variation across all groups, and, given this subspace, show how a
  Gibbs sampling algorithm can be used to estimate the posterior
  uncertainty of group-level eigenvectors and eigenvalues.  The
  implementation and utility of our model is illustrated with analyses
  of high-dimensional multivariate biological data.

@@

It has long been known that the sample covariance matrix is a poor
estimator of the population covariance matrix, particularly when the
sample size is comparable to or smaller than the number of features
($n \leq p$) \citep{Dempster1969, Stein1975}.  A host of estimators
have been devised to address this fact. These primarily include
shrinkage approaches \citep{Schafer2005, Ledoit2011} and low
rank factor models \citep{Mardia1980, Ullman2003, Fan2008}.  In many
cases, a covariance matrix is highly structured and thus can be
sparsely parameterized even if the matrix itself is not low rank
\citep{Burg1982, Williams1993} .  Along these lines, there has been a
recent interest in sparse Gaussian graphical model, in which the there
are assumed to be many zeros in the inverse covariance matrix
\citep{Meinshausen2006, Friedman2008}.

Most of these approaches assume that the goal is to estimate the
covariance matrix of a single population.  In many applications,
however, the data can be split into groups, each of which represent
samples from populations with distinct (but possibly related)
covariance matrices.  In this case, a comparative analysis of
covariance patterns across multiple groups of observations is often
warranted.  Evaluating similarities and differences between covariance
matrices can be an important complement to classical analyses which
center primarily on estimating differences in means across groups.
For instance, in case control-studies, mean level effects may be small
relative to subject variability; distributional differences between
groups may still be evident as differences in the covariances between
features.  Even when mean level differences are detectable, better
estimates of the covariability of features across groups may lead to
an improved understanding of the mechanisms underlying these apparent
mean level differences.  Further, accurate covariance estimation is an
essential part of many prediction tasks (e.g. quadratic discriminant
analysis).

Some of the earliest approaches for modeling multiple
covariance matrices focused on estimation in terms of the spectral
decompositions for these matrices.  \cite{Flury1987} developed
estimation and testing procedures for the ``common principal
components'' model, in which a set of covariance matrices were assumed
to share the same eigenvectors.  \citet{Schott1991, Schott1999}
considered cases in which only certain eigenvectors are shared across
populations, and \citet{Boik2002} described an even more general model
in which eigenvectors can be shared between all or some of the groups.
More recently, \citet{Hoff2009}, noting that eigenvectors are unlikely
to be shared exactly between groups, introduced a hierarchical model
for eigenvector shrinkage based on the matrix Bingham distribution.
There has also been a significant interest in estimating covariance
matrices using Gaussian graphical models. For Gaussian graphical
models, zeros in the precision matrix correspond to conditional
independences between pairs of features given the remaining features
\citep{Meinshausen2006}.  \citet{Witten2014} extended existing work in
this area to the multi-group setting, by pooling information about the
pattern of zeros across precision matrices.

@@

Another popular method for modeling relationships between
high-dimensional multivariate data is partial least squares regression
(PLS). This approach, which is a special case of a bilinear factor
model, involves projecting the data into a lower dimension space which
maximizes the similarity of the two groups.  This technique does not
require the data from each group to share the same feature set.  An
common variant for prediction, partial least squares discriminant
analysis (PLS-DA) is especially common in chemometrics and
bioinformatics \citep{Barker2003} .  Although closely related to the
approaches we will consider here, the primarily goal of of PLS-based
models is to create regression or discrimination models, not to
explicitly infer covariance matrices from multiple groups of data.
Nevertheless, the basic idea that data can often be well represented
on a low dimensional space is an appealing one that we leverage as
well.

In this paper we propose a multi-group covariance estimation model by
sharing information about the subspace spanned by group-level
eigenvectors.  The shared subspace assumption can be used to improve
estimates and facilitate the interpretation of differences between
covariance matrices across groups.  For each group, we assume ``the
spiked covariance model'', a well studied variant of the factor model
\citep{Johnstone2001}.  In Section \ref{sec:shared} we briefly review
the behavior of spiked covariance models for estimating a single
covariance matrix and then introduce our extension to the multi-group
setting.  In Section \ref{sec:inference} we describe an optimization
algorithm for inferring the shared subspace and then outline a MCMC
algorithm for estimating the posterior distribution of the projected
covariance matrices.  In Section \ref{sec:simulation} we
investigate the behavior of these models in simulation and then
demonstrate the utility of the model with real biological data in
Section \ref{sec:app}.

@@

%%\paragraph{The Spiked Covariance Model:}

Suppose $\bl S$ has a possibly degenerate Wishart$(\Sigma,n$)
distribution with density given by
\begin{equation} 
p(\bl S | \Sigma, n) \propto l(\Sigma:\bl S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1} \bl S/2 ) ,  
\label{eqn:lik}
\end{equation}
%
\noindent where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.  Such
a likelihood results from $\bl S$ being, for example, a residual sum
of squares matrix from a multivariate regression analysis. In this
case, $n$ is the number of independent observations minus the rank of
the design matrix.  The spiked principle components model (spiked PCA)
studied by \citet{Johnstone2001} and others assumes that
\begin{equation} 
\Sigma = \sigma^2 (  \bl U  \Lambda  \bl U^T  + \bl I )
\label{eqn:spiked}
\end{equation}
%
\noindent where for $r \ll p$, $\Lambda$ is an $r\times r$ diagonal
matrix and $\bl U \in \mathcal V_{r,p}$.  $\mathcal V_{r,p}$ is the
Stiefel manifold, which consists of the set of $p \times r$
orthonormal matrices in $\mathbb{R}^p$.  % The spiked covariance model
% is thus a low rank factor model with homoscedastic error. 
The spiked covariance formulation is appealing because it explicitly
partitions the covariance matrix into a tractable low rank ``signal''
and isotropic ``noise''.

@@

Classical results for parametric models (e.g., \citet{Schwartz1965})
imply that asymptotically in $n$ for fixed $p$, an estimator will
be consistent for a spiked population covariance as long as the
assumed number of spikes (eigenvalues larger than $\sigma^2$) is greater than or
equal to the true number.  However, when $p$ is large relative to $n$,
as is the case for the examples considered here, things are more
difficult.  Under the spiked covariance model, it has been shown that
if $p/n \rightarrow \gamma >0$ as $n\rightarrow \infty$, the $k$th largest
eigenvalue of $\bl S/(n\sigma^2)$ will converge to an upwardly biased
version of $\lambda_{k}+1$, if $\lambda_k$
is greater than $\sqrt{\gamma}$ \citep{Baik2006, Paul2007} .  This has
led several authors to suggest estimating $\Sigma$ via shrinkage of
the eigenvalues of the sample covariance matrix. In particular, in the
setting that $\sigma^2$ is known, \citet{Donoho2013} propose estimating
all eigenvalues whose sample estimates are smaller than 
$\sigma^2(1+\sqrt{\gamma})^2$ by $\sigma^2$, and shrinking the
larger eigenvalues in a way that depends on the particular loss
function being used.  These shrinkage functions are shown to be
asymptotically optimal in the $p/n\rightarrow \gamma$ setting.

@@

% Many eigenvalue shrinkage estimators can be viewed in the Bayesian
% context as derivative of a Bayes estimator under a particular prior.

%\paragraph{The Shared Subspace Model:}

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/3dplot}
        \caption{Projection in $\mathcal{R}^3$}
        \label{fig:dmelanRatio}
    \end{subfigure}
\quad
 %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter}
        \caption{$Y_kV$}
        \label{fig:dmelanPosterior}
    \end{subfigure}
\quad 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter-orth}
        \caption{$Y_kV_{\perp}$}
        \label{fig:dmelanPosterior}
    \end{subfigure}
    \caption{ Two groups of four-dimensional data (red and blue)
      projected into different subspaces.  a) To visualize $Y_k$ we
      can project the data into $R^3$.  In this illustration, the
      distributional differences between the groups are confined to a
      two-dimensional shared subspace ($VV^T$, grey plane).  b) The
      data projected onto the two-dimensional shared subspace,
      $Y_kV$, have covariances $\Psi_k$ that differ between
      groups. c) The orthogonal projection, $Y_kV_{\perp}$
      has isotropic covariance, $\sigma_k^2I$, for all groups.  }
\label{fig:shared}
\end{figure}

When estimating a single covariance matrix, the absence of prior
knowledge suggests estimators that are equivariant with respect to
rotations and scale changes.  The situation is different when we are
interested in estimating multiple covariance matrices from distinct,
but related groups.  Here, group-level equivariance to rotations is an
unreasonable assumption; both eigenvalue \emph{and} eigenvector
shrinkage can play an important role in improving covariance
estimates.

For multi-group covariance estimation, we assume that we have $K$
matrices, $Y_1, ... Y_K$, where $Y_k$ conists of where $Y_k$ consists
of $n_k$ independent rows of mean-zero normal data of dimension $p$,
typically with $n_k \ll p$.  Then, $Y_k^TY_k = \bl S_k$ has a
(degenerate) Wishart distribution as in Equation \ref{eqn:lik}.  To
improve estimation, we seek estimators for the population covariance
from one group that depend on the data from all $K$ groups, e.g. of
the form $\hat \Sigma_k(\bl S_1, ... \bl S_K)$.  To this end, we
extend the spiked covariance model to grouped data, by assuming that
the anisotropic variability from each group occurs on a common low
dimensional subspace.  Specifically, we posit that the covariance
matrix for each group can be written as

\begin{equation}
\Sigma_k = V\Psi_kV^T + \sigma^2_kI.
\label{eqn:sspsi}
\end{equation}
%
\noindent Here, the columns of $V \in \mathcal{V}_{p, s}$ determine a subspace
shared by all groups $k$ with $p \geq s$.  Throughout this paper we
will denote to the shared subspace as $VV^T \in \mathcal G_{r,p}$, where
$\mathcal G_{r,p}$, known as the Grassmann manifold, consists of all
$s$-dimensional linear subspaces of $\mathcal{R}^p$
\citep{Chikuse2012}.  Although $V$ is only identifiable up to right
rotations, the matrix $VV^T$, which defines the plane of variation shared by all
groups, is identifiable.  Later, to emphasize the connection to the
spiked PCA model (\ref{eqn:spiked}), we will write $\Psi_k$ in terms
of its eigendecomposition, $\Psi_k = O_k\Lambda_kO_k$, where $O_k$ are
eigenvectors and $\Lambda_k$ are the eigenvalues of $\Psi_k$ (see
Section \ref{sec:bayes}).

Note that for the shared subspace model,
$V^T\Sigma_kV = \Psi_k + \sigma_k^2I$ is an anisotropic
$s$-dimensional covariance matrix for the projected data, $Y_kV$.
In contrast, the data projected onto the orthogonal space,
$Y_kV_{\perp}$, is isotropic, and assumed to be small, for all
groups.  In Figure \ref{fig:shared} we provide a simple illustration
using simulated $4$-dimensional data from two groups.  In this example, the
differences in distribution between the groups of data can be
expressed on a two dimensional subspace spanned by the columns of
$V \in \mathcal{V}_{4, 2}$.  Differences in the correlations between
the two groups manifest themselves on this shared subspace, whereas
only the magnitude of the isotropic variability can differ between groups
on the orthogonal space.  Thus, the the shared subspace model can be
viewed as a covariance partition model, where one partition includes
the anisotropic variability from all groups and the other partition is
constrained to the isotropic variability from each group.  This
isotropic variability is often characterized as measurement noise.

@@

\section{Empirical Bayes Inference}
\label{sec:inference}

In this section we outline our empirical Bayesian approach for estimating
covariance parameters and their associated uncertainty.  
Uncertainty quantification is particularly important in the $n \ll p$
setting where seemingly large differences in the point estimates of
the low rank covariance matrices across groups may not actually
reflect significant differences.  Although uncertainty quantification
for all parameters is desirable, for the size of the problems we are
generally interested in, we found MCMC inference for the subspace
$VV^T$ to be intractable.  Our optimization based approached is
significantly faster and works well for high-dimensional datasets.  As
such, we start, in Section \ref{sec:em}, by outlining an
expectation-maximization algorithm for estimating the shared
subspace. In Section \ref{sec:bayes} we describe a Gibbs sampler for
posterior inference of $\Psi_k$ given the common subspace.  These two
algorithms can be used in conjunction to first identify and evaluate a
shared subspace and then to quantify the uncertainty associated with
the projected covariance matrices.

TODO: emphasize 1) justification 2) objective of \emph{identifying} this
subspace when p much greater than n.

@@

In this section we describe a maximum marginal likelihood procedure for
estimating the shared subspace, $VV^T$ based on the
expectation-maximization algorithm.  The full likelihood
for the shared subspace model can be written as

\begin{align}
\nonumber p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}\mathbf{S}_k/2)  \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\Psi_kV^T +
  I))^{-1}\mathbf{S}_k/2) \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\Psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]\mathbf{S}_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]\mathbf{S}_k/2) ,
\end{align}
%
\noindent where we define $M_k^{-1} = (\Psi_k + I) ^{-1}/\sigma_k^2$.  The log-likelihood in
$V$ (up to an additive constant) is

\begin{align}
\nonumber l(V) &= \sum_k \tr\left(-VM_k^{-1}V^T +
       VV^T/\sigma^2_k\right)\mathbf{S}_k/2)\\
&= \sum_k \frac{1}{2}\tr\left(-M_k^{-1}V^T \mathbf{S}_kV\right) + \frac{1}{2\sigma_k^2}tr\left(VV^T \mathbf{S}_k\right).
\end{align}

@@

\noindent which is a Wishart($V^T\mathbf{S}_kV$, $n_k$) distribution.  The
conditional posterior distribution of $\frac{1}{\sigma_k^2}$ is simply

@@

\noindent which is a gamma($n_k(p-r)/2$,
$\frac{1}{2}tr((I-VV^T)\mathbf{S}_k)$) distribution.  Based on these
results the following EM algorithm can be used.

\begin{enumerate}
\item For each $k$, compute relevant conditional expectations:
\begin{enumerate}
\item $E[M_k^{-1} | V_{(t-1)}] = n_k(V_{(t-1)}^T \mathbf{S}_kV_{(t-1)})^{-1}$
\item $E[\frac{1}{2\sigma_k^2}|V_{(t-1)}] = \frac{n_k(p-r)}{tr((I-V_{(t-1)}V_{(t-1)}^T)S_k)}$
\end{enumerate}
\item Compute $V_{t}$ = $\underset{V}{\text{arg } \text{max}}  \sum_k tr\left(-VE[M^{-1}|V_{(t-1)}]^T +
       E[\frac{1}{2\sigma_k^2}|V_{(t-1)}]VV^T\right)\mathbf{S}_k/2)$ 
\end{enumerate}

@@

\paragraph{Evaluating Goodness of Fit:}
